---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# lemna

<!-- badges: start -->
<!-- badges: end -->

## Overview


*lemna* is a software package for the language [*R*](https://www.r-project.org/).
It implements model equations and default parameters to simulate the
toxicokinetic-toxicodynamic (TKTD) model of the *Lemna* aquatic plant.
*Lemna* is a standard test macrophyte used in ecotox effect studies. The model
was described and published by the *SETAC Europe Interest Group Effect Modeling*
(Klein *et al.* 2021). It is a refined description of the *Lemna* TKTD model
published by Schmitt *et al.* (2013).


## Installation

```{r, eval=FALSE}
# install directly from CRAN
install.packages("lemna")

# install latest development version from GitHub
devtools::install_github("nkehrein/lemna")
```

## Usage

The lemna package provides model equations and some useful helpers to simulate
the growth of *Lemna* (duckweed) aquatic plant populations. A simulation is run
by using the `lemna()` function. The required scenario data are either supplied
individually on function call or are passed as a pre-defined scenario object,
such as the `focusd1` sample scenario:

```{r example}
library(lemna)

# simulate sample scenario
result <- lemna(focusd1)
head(result)

plot(result)
```

## Learning lemna

The package contains two vignettes that may help you getting started:

 - [Introduction to the Lemna package](https://nkehrein.github.io/lemna_web/lemna-introduction.html)<br />
   A general *Tutorial* and guide to the package functions
 - [Lemna model verification](https://nkehrein.github.io/lemna_web/lemna-verification.html)<br />
   A verification of the model implementation against results of the Schmitt
   *et al.* implementation. Contains advanced workflows of package features.

## License

The package and its source code is free and open-source software available under
the MIT license.

## Issues

If you find any issues or bugs within the package, please create a
[new issue](https://github.com/nkehrein/lemna/issues) on GitHub.

## References

- Klein J, Cedergreen N, Heine S, Reichenberger S, Rendal C,
  Schmitt W, Hommen U. 2021. Refined description of the *Lemna* TKTD growth model
  based on *Schmitt et al.* (2013) – equation system and default parameters.
  Report of the working group *Lemna* of the SETAC Europe Interest Group Effect
  Modeling. Version 1, uploaded on 22. Sept. 2021.
  https://www.setac.org/group/SEIGEffectModeling
- Schmitt W, Bruns E, Dollinger M, Sowig P. 2013. Mechanistic TK/TD-model
  simulating the effect of growth inhibitors on *Lemna* populations. Ecol Model
  255, pp. 1-10. DOI: [10.1016/j.ecolmodel.2013.01.017](https://doi.org/10.1016/j.ecolmodel.2013.01.017)
