---
title: "Introduction to the Lemna package"
author: "Nils Kehrein"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Introduction to the package}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7.2,
  fig.height = 5,
  fig.align = "center"
)

# do not build vignette on package checks
is_check <- ("CheckExEnv" %in% search()) || any(c("_R_CHECK_TIMINGS_",
             "_R_CHECK_LICENSE_") %in% names(Sys.getenv()))
knitr::opts_chunk$set(eval = !is_check)
```


The *lemna* package provides model equations and some useful helpers to
simulate the growth of *Lemna* (duckweed) aquatic plant populations. 
*Lemna* is a standard test macrophyte used in ecotox effect studies. The
model was described and published by the SETAC Europe Interest Group Effect
Modeling (Klein *et al.* 2021).

The model's main state variable is biomass, or `BM` for short, of the simulated
*Lemna* population. Growth of *Lemna* is influenced by environmental variables
such as temperature, irradiation, nutrient concentrations, population density,
and toxicant concentration in the surrounding medium. To consider the influence
of toxicants on the plants, a one-compartment model was assumed by the authors for the
mass-balance of internal toxicant mass. The total amount of internal toxicant
mass is represented by state-variable `M_int`. The
combination of state variables `BM` and `M_int` fully describe the state of the
model system at any point in time.

To simulate a *Lemna* population, one has to define a *scenario* that consists
of the following data:

- Initial system state
- A time period to simulate
- Model parameters
- Environmental variables

How these scenario elements are represented and which values are chosen depends
on what one would like to achieve. Simulating the growth of *Lemna* in a controlled
lab environment will likely require different inputs than *Lemna* growing in an outdoor
water body, for example.



## Core functions

To make functions and sample datasets of the *lemna* package available in your
R workspace, load the library first:

```{r setup}
library(lemna)
```

The package function `param_defaults()` provides a list with all suggested 
default parameters. Some parameter values will be missing, i.e.
set to `NA`, because they are substance specific and default values would not
be meaningful for these:

```{r}
# get list of default parameters
params <- param_defaults()
params$k_photo_max
params$EC50_int # substance specific

# get default parameters and set a custom parameter value
myparam <- param_defaults(c(EC50_int = 42))
myparam$EC50_int
```


The growth of a *Lemna* population is simulated using the `lemna()` function. 
The required scenario data are either supplied individually on function call
or are passed as a pre-defined scenario object, such as the `metsulfuron` sample
scenario:

```{r}
lemna(metsulfuron)
```

`lemna()` returns a table which describes the change of state variables over time.
In addition, some supporting derived variables such as internal toxicant concentration
(`C_int`) and the number of fronds (`FrondNo`) will be returned by default.

A visual description of the simulated scenario and its results can be created
by running the `plot()` function. The `plot()` function requires a simulation result
as its first argument:

```{r}
plot(lemna(metsulfuron))
```


The effect of the toxicant on the *Lemna* population can be calculated using
the `effect()` function. It requires scenario data the same way as `lemna()`
does. For the sample `metsulfuron` scenario, the effects of the toxicant
are as follows:

```{r}
effect(metsulfuron)
```

In this scenario, exposure to the toxicant resulted in an 93% decrease of
population size (`BM`) and a 46% decrease in average growth rate (`r`) until the
end of the simulation. Effects are always calculated relative to an identical
control scenario which contains no toxicant exposure.

For more information on the `metsulfuron` sample scenario, please refer to the
help files:

```{r eval=FALSE}
?metsulfuron
```



## Tutorial
### Simulate the *Lemna* growth model

To simulate a *Lemna* population, one has to pass the four mandatory scenario
elements to the `lemna()` function:

```{r}
# initial state of the model system: 1.0 g dw biomass, 0.0 ug/m2 internal toxicant
myinit <- c(BM=1, M_int=0)
# simulated period and output time points: each day for 7 days
mytimes <- 0:7
# default model parameters + substance specific values
myparam <- param_defaults(c(
  EC50_int = 4.16,
  b = 0.3,
  P = 0.0054
))
# constant environmental conditions, including exposure
myenvir <- list(
  tmp = 18,    # 18 °C ambient temperature
  irr = 15000, # 15,000  kJ m-2 d-1 irradiance
  P = 0.3,     # 0.3 mg L-1 Phosphorus concentration
  N = 0.6,     # 0.6 mg L-1 Nitrogen concentration
  conc = 1     # 1 ug/L toxicant concentration
)

lemna(
  init = myinit,
  times = mytimes,
  param = myparam,
  envir = myenvir
)
```

The `init` argument controls at which system state the simulation starts. The 
`times` argument defines the length of the simulated period and for which
time points results are returned. The temporal resolution of results can be increased
by specifying additional output times:

```{r}
simresult <- lemna(
  init = myinit,
  times = seq(0, 7, 0.1), # a step length of 0.1 days = ~2 hours
  param = myparam,
  envir = myenvir
)
tail(simresult)
```

The resulting table now contains ten times as much rows because we decreased the
step length by a factor of ten but simulated the same period, i.e. seven days.
It can be observed that the state-variables differ slightly at the end of the
simulation although the scenarios were otherwise identical. The differences
originate from small numerical errors introduced by the solver of the model's
Ordinary Differential Equations (ODE). The step-length in time can have influence on
the precision of simulation results. To decrease the solver's step length
without increasing the number of result time points, make use of the optional
argument `hmax`. The smaller `hmax`, the more precise the results:

```{r}
# hmax=0.01 forces a maximum step length of 0.01 days = ~15 minutes
lemna(myinit, mytimes, myparam, myenvir, hmax = 0.01)
```

By default, simulation results contain supporting variables such as
internal toxicant concentration and total frond number. These are calculated
from simulation results and model parameters for reasons of convenience. If these
variables are not required, they can be disabled by setting the optional argument
`nout = 0`:

```{r}
lemna(myinit, mytimes, myparam, myenvir, nout = 0)
```




### Using environmental time-series

The previous examples mostly assumed that environmental variables stay constant in time.
To simulate a scenario with changing environmental variables, such as a temperature
curve or exposure pattern, one has to define or load a data time-series. The model
accepts time-series for all environmental variables, i.e. exposure concentration,
temperature, irradiation, phosphorus concentration, and nitrogen concentration.

Within the scope of this package, time-series are represented by a 
`data.frame` containing exactly two numerical columns: the first column for time, the second
for the variable's value. The column names are irrelevant but sensible names may
help documenting the data. As an example, the `metsulfuron` sample scenario
contains a step-function as its exposure time-series: seven days of
1 ug/L *metsulfuron-methyl* starting at time point zero (`0.0`), followed
by seven days of recovery (no exposure).

```{r}
metsulfuron$envir$conc
```

Time points of the time-series and time points processed by the ODE solver
may not always match. To derive environmental variable values which are not
explicitly part of the time-series, variable values are interpolated with a linear
function.
If the time-series does not cover the full simulation period, the closest
value from the time-series is used. In the case of the `metsulfuron` sample
scenario, the step function will effectively extend to infinity, i.e. any time
point before day `7.0` will have 1 ug/L of exposure and any time point after
`7.01` will have no exposure.

As an example, we will modify the `metsulfuron` sample scenario to use an
exposure time-series that declines linearly between start and day seven:

```{r}
# define start and end points for the exposure series, the values
# in between will be interpolated
myexpo <- data.frame(time=c(0, 7), conc=c(1, 0))
# modify the sample scenario's exposure series
myenvir <- metsulfuron$envir
myenvir$conc <- myexpo

# simulate the sample scenario with modified environmental variables
plot(lemna(metsulfuron, envir=myenvir))
```

Time-series and `data.frame` objects can be stored conveniently as `.csv` files
which can be created and edited by common spreadsheet programs such as *Microsoft
Excel*. Be aware that the separator character used by *R* and your spreadsheet
program may differ depending on your computer's locale settings.

```{r, fig.width = 5, fig.height = 4,}
set.seed(23)
# define a random time-series, values will be uniformly distributed between
# the values 0.1 and 3.0, e.g to represent an exposure time-series
myexpo <- data.frame(time = 0:14,
                     conc = round(runif(15, 0.1, 3.0), 1))
# plot the time-series
plot(myexpo, main="Random exposure time-series")
lines(myexpo)
# write data to .csv file in working directory
write.csv(myexpo, file="random_series.csv", row.names=FALSE)
# write data using semicolons as separating character
write.csv2(myexpo, file="random_series2.csv", row.names=FALSE)

# read file from working directory
myimport <- read.csv(file="random_series.csv")
# check that written and read data are identical
myexpo$conc == myimport$conc
```

Time-series can be imported manually as in the previous example or they can be
imported automatically by the `lemna()` function for convenience. If an
environmental variable is set to a string, it will be interpreted as a file path
and `lemna()` will try to import the time-series using `read.csv()`:

```{r}
# automatically load the exposure time-series from a file
myenvir <- metsulfuron$envir
myenvir$conc <- "random_series.csv"

# simulate the sample scenario with the exposure series loaded from a .csv file
plot(lemna(metsulfuron, envir=myenvir), legend=FALSE)
```

```{r, include = FALSE}
# clean up vignette directory
file.remove("random_series.csv")
file.remove("random_series2.csv")
```


For a more complex scenario that uses hourly and daily time-series of
exposure and temperature/irradiance, respectively, please have a look at
e.g. the `focusd1` scenario:

```{r, eval=FALSE}
myenvir <- focusd1$envir
myenvir$conc
myenvir$tmp
myenvir$irr
```



### Using simulation results

Simulation results are returned as a table, i.e. a `data.frame` object. The table
will contain the state variables biomass (`BM`) and internal toxicant mass (`M_int`)
for each requested output time point. The table may also contain additional
columns for other supporting variables. The
data can be processed like any other dataset in *R* to e.g. create plots, derive
other values, or to perform statistical tests:

```{r}
myresult <- lemna(focusd1)
head(myresult)
```

To get an initial impression of a scenario and its results, simply pass the
simulation result to the `plot()` function:

```{r}
plot(myresult)
```


As an example, we will analyze if and how the internal toxicant concentration
(`C_int`) correlates with the internal toxicant mass (`M_int`):

```{r}
summary(lm(C_int ~ M_int, myresult))
```

The linear model indicates a strong correlation of internal toxicant mass
and concentration which intuitively makes sense. The correlation is not a
100% because biomass is a confounding factor in the model equations.



### Derive effect endpoints

To quantify the influence a toxicant exerts on a *Lemna* population, use the
`effect()` function. It works similar to `lemna()` and accepts the same
arguments in order to specify a scenario:

```{r}
# calculate effects on biomass in sample scenario
effect(metsulfuron)
```

The return values describe the effect in percent (%) on the respective effect
endpoint. Effects are calculated relative to a control scenario which exhibits
no exposure. By default, the effect refers to the reduction in biomass (`BM`) or
average growth rate (`r`) at the end of the simulation. In the example above,
biomass was reduced by 93% and the growth rate was reduced by 46% in the *Lemna*
population due to exposure to the toxicant.

If a scenario covers a long time period but effects are desired for an earlier
time point, the scenario can be cut short by using the `duration` argument.
If `duration` is set, the scenario will be clipped to the time period from `t0`
to `t0 + duration`:

```{r}
# calculate effects on biomass after 7 days, instead of 14
effect(metsulfuron, duration=7)
```

In this example, the effect on biomass is smaller after 7 days compared to the
effects after 14 days. However, the average growth rate experienced a strong decrease
from 46 to 71%.



### Create scenario objects

A *Lemna* growth scenario consists of the following four mandatory scenario
elements: model parameters, environmental variables, initial state, and output times.
The elements can be passed to `lemna()` and `effect()` separately or they can be
combined to a compact scenario object. All sample scenarios which were
used in this tutorial are scenario objects:

```{r, eval=FALSE}
# list properties of the sample scenario object
metsulfuron
```

Scenario objects are basically just a base *R* `list` object with some additional
metadata. If correctly defined, scenario objects fully describe a scenario and
can be passed to e.g. `lemna()` without additional arguments. It is, however,
possible to override a scenario object's data by passing an alternative
dataset:

```{r}
# custom output times and time period:
# four days with a 12 hour time step
mytimes <- seq(0, 4, 0.5)

# simulate sample scenario with custom output times & period
lemna(metsulfuron, times=mytimes)
```

A custom scenario object can be created by passing the scenario elements to
`new_lemna_scenario()`:

```{r}
myscenario <- new_lemna_scenario(
  init = c(BM=1, M_int=0),
  times = 0:7,
  param = param_defaults(c(EC50_int = 4.16, b = 0.3, P = 0.0054)),
  envir = list(
    tmp = 18,    # 18 °C ambient temperature
    irr = 15000, # 15,000  kJ m-2 d-1 irradiance
    P = 0.3,     # 0.3 mg L-1 Phosphorus concentration
    N = 0.6,     # 0.6 mg L-1 Nitrogen concentration
    conc = 1     # 1 ug/L toxicant concentration
  )
)
lemna(myscenario)
```



### Speed up simulations with compiled code

The *Lemna* growth model is simulated by default using model equations
implemented in pure *R*. In case many simulations have to be conducted
or the time required to get results becomes an issue, the compiled code
feature can be used. The *lemna* package provides an alternative implementation
of the *Klein et al.* model equations using *C* code. The *C* code
executes significantly faster than the pure *R* alternative.

```{r}
# use model implemented in pure R
tail(lemna(metsulfuron, ode_mode="r"), n = 1)

# use model implemented in C
tail(lemna(metsulfuron, ode_mode="c"), n = 1)
```


Simulation results of *R* and *C* code will be identical as far as numerical
precision allows. The speed increase of using *C* will range from a factor
of 3 to 5 for short scenarios and up to 50+ for longer scenarios:

```{r}
# Benchmark the shorter metsulfuron scenario
microbenchmark::microbenchmark(
  lemna(metsulfuron, ode_mode="r"),
  lemna(metsulfuron, ode_mode="c")
)

# Benchmark the more complex and longer focusd1 scenario
microbenchmark::microbenchmark(
  lemna(focusd1, ode_mode="r"),
  lemna(focusd1, ode_mode="c"),
  times = 10
)
```

There is however a small disadvantage to using the *C* model: if there are
any issues stemming from, for example, invalid parameters, the error messages
raised by the *C* code might be less descriptive than those from *R*.
On the other hand, the *C* code can output on demand almost all intermediary model
variables which can support debugging and model understanding:

```{r}
# simulate and request all additional output variables
lemna(metsulfuron, ode_mode="c", nout=18)
```









## References
- Klein J., Cedergreen N., Heine S., Reichenberger S., Rendal C.,
  Schmitt W., Hommen U., 2021: Refined description of the *Lemna* TKTD growth model
  based on *Schmitt et al.* (2013) – equation system and default parameters.
  Report of the working group *Lemna* of the SETAC Europe Interest Group Effect
  Modeling. Version 1, uploaded on 22. Sept. 2021.
  https://www.setac.org/group/SEIGEffectModeling
